---
title: "Genomic Imprinting Analysis in *Mimulus nudatus* Hybrid Endosperm"
author: "Elen Oneal"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(here)
library(knitr)
library(rmarkdown)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggvenn)
library(edgeR)
library(limma)

source(here('scripts', 'R', 'edger_preanalysis.R'))
source(here('scripts', 'R', 'plot_mds.R'))
source(here('scripts', 'R', 'merge_snpcounts.R'))
source(here('scripts', 'R', 'snpcounts_by_parent.R'))
source(here('scripts', 'R', 'filter_snpcounts.R'))
source(here('scripts', 'R', 'imprinting_analysis.R'))
source(here('scripts', 'R', 'plot_imprinted_genes.R'))
source(here('scripts', 'R', 'get_megs_pegs.R'))
```

## Introduction

Genomic imprinting is an epigenetic phenomenon in which gene expression 
is biased toward one parental allele. In flowering plants, imprinting is 
largely restricted to the endosperm, a triploid nutritive tissue that 
supports embryo development. Because the endosperm is formed by fusion of 
the diploid maternal central cell with a haploid paternal gamete, most 
endosperm-expressed genes conform to an expected 2:1 maternal:paternal 
expression ratio. Imprinted genes deviate from this expectation. 
Maternally expressed genes (MEGs) show maternal bias, while paternally 
expressed genes (PEGs) show paternal bias.

Imprinting is theorized to result from genomic conflict between maternal 
and paternal genomes over resource allocation to offspring. Disruption in imprinting and the subsequent disruption in endosperm tissue development can lead to hybrid seed inviability, and is thought to be an important factor in the emergence of reproductive isolation (speciation) flowering plants. If so, then imprinted genes may play a causal role in the emergence of barriers to hybridization between closely related species. This project is part of a larger research study that combines rna sequencing of endosperm transcriptomes with QTL mapping in reciprocal hybrid backcross populations to identify and characterize genomic loci contributing to hybrid seed inviability.

Here we characterize genomic imprinting in hybrid endosperm from reciprocal 
crosses within the *Mimulus guttatus* species complex, focusing on crosses 
between *M. nudatus* inbred lines DHRO22 and GMR2. We use allele-specific 
expression (ASE) analysis to identify MEGs and PEGs, leveraging 
parent-of-origin SNPs identified from parental DNA-seq data. 

## Study Design

Reciprocal crosses were performed between two *M. nudatus* inbred lines:

- **DHRO22 × GMR2** — DHRO22 as maternal parent (3 replicates)
- **GMR2 × DHRO22** — GMR2 as maternal parent (4 replicates)

RNA-seq libraries were generated from hybrid endosperm tissue. Reads were 
aligned to the *M. guttatus* IM767 v2 reference genome using STAR. 
Allele-specific read counts were quantified at parent-of-origin SNPs 
identified from parental DNA-seq using BWA and bcftools. Imprinting was 
tested using a likelihood ratio test in edgeR, with cross direction as a 
covariate to control for maternal effects.

In addition to these samples, we also sequenced hybrid endosperms in multiple M. guttatus crosses. Finally, we sequenced whole fruits of 2 *M. guttatus* and 3 *M. nudatus* fruits. These samples are included for visualization purposes only. 

## MDS Analysis

We first examined sample clustering using multidimensional scaling (MDS) 
to assess overall sample quality and verify that samples cluster by 
species and cross direction.  

The first MDS plot includes all samples RNA sequenced in this study.

```{r load_endosperm_count_object, echo=FALSE, eval=TRUE}

# Load featureCounts object
fc <- readRDS(here('config', 'counts_object'))

# Build gene length table
gl <- data.frame(fc$annotation$GeneID, fc$annotation$Length)
colnames(gl) <- c('gene_name', 'length')
gl$gene_name <- gsub('.v2.1', '', gl$gene_name)

# Build count matrix
gc <- as.data.frame(fc$counts)
colnames(gc) <- gsub('.2.Aligned.sortedByCoord.out.bam', '', colnames(gc))
gc$gene_name <- gsub('.v2.1', '', rownames(gc))
rownames(gc) <- c()

# Reorder columns
gc <- gc %>% select(
  matches('_RNA'), matches('CSS4x'), matches('IM767x'),
  matches('DHR14x'), matches('SODx'), matches('DHRO22x'),
  matches('GMR2x'), everything()
)

# Build samples data frame for new edger_preanalysis signature
samples_df <- data.frame(
  sample = colnames(gc)[colnames(gc) != "gene_name"],
  group  = c(1,1,1,1,2,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,7)
)

e <- edger_preanalysis(gc, samples_df, 1, 5, 'wholeseed and endosperm')
x <- e$y


# MDS plot
group_labels <- c(
  '1' = 'Wholeseed',
  '2' = 'CSS4xIM767',
  '3' = 'IM767xCSS4',
  '4' = 'DHR14xSOD',
  '5' = 'SODxDHR14',
  '6' = 'DHRO22xGMR2',
  '7' = 'GMR2xDHRO22'
)

plot_mds_gg(x,
            color_var = 'group',
            labels    = group_labels,
            title     = 'MDS: wholeseed and endosperm')
```

We then restrict our analysis to the *M. nudatus* hybrid endosperm. 

```{r nudatus_endosperm_expressed_genes, echo=FALSE, eval=TRUE}

# Subset to M. nudatus endosperm samples only
ngc <- gc %>% select(matches('DHRO22x'), matches('GMR2x'), gene_name)

# Build samples data frame
nudatus_df <- data.frame(
  sample = colnames(ngc)[colnames(ngc) != "gene_name"],
  group  = factor(c(1, 1, 1, 2, 2, 2, 2))
)

# Run edgeR pre-analysis
ne <- edger_preanalysis(ngc, nudatus_df, 1, 2, 'nudatus endosperm')
nx <- ne$y

# Smear plot inline
#plotSmear(ne$qlf, main = "Smear plot: M. nudatus endosperm")
#abline(h = c(-2, 2), col = "gray", lty = 2)

# MDS plot
group_labels <- c('1' = 'DHRO22xGMR2', '2' = 'GMR2xDHRO22')
mds_nudatus <- plot_mds_gg(nx,
                            color_var = 'group',
                            labels    = group_labels,
                            title     = 'MDS: M. nudatus hybrid endosperm')
mds_nudatus

# Compute minimum CPM threshold from nudatus library sizes
# CPM equivalent of 10 counts in the smallest nudatus library
min.cpm.nudatus <- 10 / (min(nx$samples$lib.size) / 1000000)
message(sprintf("Min CPM threshold: %.2f (= 10 counts in smallest library of %.1fM reads)",min.cpm.nudatus, min(nx$samples$lib.size)/1000000))

# Filter for genes expressed above threshold in at least 2 samples
keep <- rowSums(cpm(nx) > min.cpm.nudatus) >= 2
nud.gene.df   <- ne$df_kept[keep, ]
nud.gene.list <- as.list(rownames(nud.gene.df))

message(sprintf("Genes expressed in M. nudatus endosperm: %d", 
                length(nud.gene.list)))

# Write raw counts for expressed genes
write.table(
  data.frame(nud.gene.df, gene_name = rownames(nud.gene.df)),
  here('results', 'nudatus_endosperm_raw_counts.txt'),
  quote = FALSE, row.names = FALSE, sep = '\t'
)
```

We next examined the overlap between genes expressed in *M. nudatus* 
endosperm and genes with informative parent-of-origin SNPs, since only 
genes in both sets can be tested for imprinting.

```{r nudatus_gene_snps, echo=FALSE, eval=TRUE}
nudatus.snps <- read.table(here('config', 'dhro22_gmr2_coding_sequence_snps.bed'),
                           h = TRUE, sep = '\t')
x <- list('Expressed genes' = nud.gene.df$gene_name,
          'Genes with parental SNPs' = nudatus.snps$Gene)
ggvenn(x, text_size = 8, show_percentage = FALSE)
```

Load files with gene information: mimulus/arabidopsis orthologs, synonyms for arabidopsis genes, defline for arabidopsis genes, and coordinates for genes in IM767 v2 reference genome.

```{r,get_gene_info,include=FALSE,eval=TRUE}
mim_arab_orthologs <- readRDS(here('config','mimulus_arabidopsis_orthologs'))
arab_syn <- readRDS(here('config','arabidopsis_synonyms'))
arab_defline <- readRDS(here('config','arabidopsis_defline'))
gene.coord <- read.table(here('config','im767.gene_coordinates.txt'),h=T)
```

Load allelic counts for each sample in the hybrid endosperm RNA sequencing experiment. Perform imprinting analysis using edgeR. We use edgeR to fit a negative binomial GLM incorporating both read source (maternal vs paternal) and cross direction as covariates. A likelihood ratio test (LRT) identified genes with significant parent-of-origin biased expression while controlling for cross direction effects, with FDR correction using the Benjamini-Hochberg method.

```{r nudatus_imprinting, echo=FALSE, eval=TRUE,message=TRUE}
fdr_cutoff <- 0.05
cross        <- 'DHRO22xGMR2'
reversecross <- 'GMR2xDHRO22'
species      <- 'M. nudatus'
# Load and merge per-sample allele count files
nc <- multiMerge(here('results'), 
                 '.nudatus.allelecounts.txt', 
                 'dhro22', 'gmr2')

# Assign maternal/paternal labels based on cross direction
nc <- prepare_snpcounts(nc, 'DHRO22xGMR2', 'dhro22', 'gmr2')

# Write raw SNP counts
write.table(
  data.frame(nc, gene_name = row.names(nc)),
  here('results', paste0('nudatus_raw_snpcounts.txt')),
  quote = FALSE, row.names = FALSE, sep = '\t'
)

# Run parent-specific expression analysis
# cpm.filter=2, n.cpm.filter=6 (must exceed threshold in all samples)
invisible(capture.output(
  pse.nud <- parent_specific_expression(nc, 2, 6, 'DHRO22xGMR2', 'DHRO22', 'GMR2', fdr_cutoff)
))

normalized_nudatus_counts  <- pse.nud[[1]]
pse.nud.degs               <- pse.nud[[2]]
pse.nud.edgeR.lrt          <- pse.nud[[3]]
pse.nud.edgeR.fit          <- pse.nud[[4]]
pse.nud.edgeR              <- pse.nud[[5]]
pse.nud.significant_genes  <- pse.nud[[6]]
nud.imp.exp                <- pse.nud[[7]]

# Require minimum 3 informative SNP reads summed across replicates
# in both cross directions to ensure reliable maternal proportion estimates
nud.imp.exp <- nud.imp.exp %>% filter(total1 >= 10 & total2 >=10)
pse.nud.significant_genes <- pse.nud.significant_genes %>% 
  filter(gene_name %in% nud.imp.exp$gene_name)

# Imprinting scatter plot
nud.plot <- imprinting_plot(nud.imp.exp, pse.nud.significant_genes, 
                            cross, reversecross, species)
nud.plot

ggsave(here('results', 'figures', paste0('Allelic_Imbalance_', cross, '.pdf')),
       plot = nud.plot, width = 8, height = 6)

# Maternal proportion histogram
maternal_histogram(nud.imp.exp, species)

# Get MEG and PEG lists
l <- get_imprinted_list(nud.imp.exp, cross, reversecross, 
                        pse.nud.significant_genes, pse.nud.edgeR.lrt, fdr_cutoff)

nud.imp.exp  <- l[[1]]
nud.megs     <- l[[2]]
nud.pegs     <- l[[3]]
nud.results  <- l[[4]]

message(sprintf("MEGs: %d", nrow(nud.megs)))
message(sprintf("PEGs: %d", nrow(nud.pegs)))
message(sprintf("Mean maternal proportion MEGs: %.3f", mean(nud.megs$mean_maternal)))
message(sprintf("Mean maternal proportion PEGs: %.3f", mean(nud.pegs$mean_maternal)))

```

```{r nudatus_imprinting_summary, echo=FALSE, eval=TRUE}

# Summary statistics
n_expressed   <- nrow(nud.imp.exp)
n_megs        <- nrow(nud.megs)
n_pegs        <- nrow(nud.pegs)
mean_meg_mat  <- mean(nud.megs$mean_maternal)
mean_peg_mat  <- mean(nud.pegs$mean_maternal)
mean_peg_pat  <- 1 - mean_peg_mat

# Print as a clean table
summary_df <- data.frame(
  Metric = c(
    'Genes with informative SNPs',
    'Maternally expressed genes (MEGs)',
    'Paternally expressed genes (PEGs)',
    'Mean maternal proportion in MEGs',
    'Mean paternal proportion in PEGs'
  ),
  Value = c(
    n_expressed,
    n_megs,
    n_pegs,
    sprintf('%.1f%%', mean_meg_mat * 100),
    sprintf('%.1f%%', mean_peg_pat * 100)
  )
)

knitr::kable(summary_df, 
             col.names = c('Metric', 'Value'),
             align = c('l', 'r'),
             caption = paste0('Summary of imprinting analysis in ',
                             '*M. nudatus*', ' hybrid endosperm'))
```

```{r nudatus_annotate_results, echo=FALSE, eval=TRUE}
# Annotate with Arabidopsis orthologs and gene coordinates
nud.megs <- list(nud.megs, nud.results, mim_arab_orthologs, gene.coord) %>%
  Reduce(function(d1, d2) left_join(d1, d2, by = 'gene_name'), .)
nud.megs <- list(nud.megs, arab_syn, arab_defline) %>%
  Reduce(function(d1, d2) left_join(d1, d2, by = 'arabidopsis_name'), .) %>%
dplyr::select(gene_name, contig, start,
              all_of(paste0(cross, '_prop_maternal')),
              all_of(paste0(reversecross, '_prop_maternal')),
              mean_maternal, logFC, logCPM, LR, PValue, P_adj,
              arabidopsis_name, synonym, defline)

nud.pegs <- list(nud.pegs, nud.results, mim_arab_orthologs, gene.coord) %>%
  Reduce(function(d1, d2) left_join(d1, d2, by = 'gene_name'), .)
nud.pegs <- list(nud.pegs, arab_syn, arab_defline) %>%
  Reduce(function(d1, d2) left_join(d1, d2, by = 'arabidopsis_name'), .) %>%
 dplyr::select(gene_name, contig, start,
              all_of(paste0(cross, '_prop_maternal')),
              all_of(paste0(reversecross, '_prop_maternal')),
              mean_maternal, logFC, logCPM, LR, PValue, P_adj,
              arabidopsis_name, synonym, defline)%>%
  unique()

# Write gene lists
write.table(nud.megs %>% select(gene_name),
            here('results', paste0('MEGs_', cross, '_list.txt')),
            row.names = FALSE, sep = '\t', col.names = TRUE, quote = FALSE)

write.table(nud.pegs %>% select(gene_name),
            here('results', paste0('PEGs_', cross, '_list.txt')),
            row.names = FALSE, sep = '\t', col.names = TRUE, quote = FALSE)

# Write annotated gene lists
write.table(nud.megs,
            here('results', paste0('MEGs_', cross, '_annotated.txt')),
            row.names = FALSE, quote = FALSE, col.names = TRUE, sep = '\t')

write.table(nud.pegs,
            here('results', paste0('PEGs_', cross, '_annotated.txt')),
            row.names = FALSE, quote = FALSE, col.names = TRUE, sep = '\t')
```


